---
title: "BBets"
author: "Tiffany Alvarado"
output: html_document
date: "2024-11-5"
---

This assignment is designed to:
1. use ChatGPT to develop Rmarkdown code
2. illustrate the use of simulation to estimate the probability of 2 or more people having the same birthday (month, day) from a random sample n 

This assignment has three questions:  

use the prompt:

write rmarkdown code to find the probability that 2 or more people in a class of size 35 have the same birthday (month, day) by simulation. Print this probability. Repeat this process 100 times, save each probability in a vector, probs. Print the mean and standard deviation of probs.  Plot a histogram of this probs. Get a 95 percent confidence interval of the mean of probs.  Print the amount of money Dr. J bets at least 2 students in the class have the same birthday (month,day) How much should Dr. J have to bet against the student's $1 to make this a fair bet.
  

Then answer the following 3 questions:

1. in a class of 35 students, Dr. J bets $1 against student's $1  that at least 2 students have the same birthday, (month, day).  
a)Find the probability that at least two student have the same birthday.  
b)If 26 students in the class take the bet how much is Dr. J expected to win/lose.   
c)How much should Dr. J wager against the student's $1 to make this a fair bet?  Vary the payoff (Dr. J should pay more that $1 against the student's $1 bet) to make the mean probability 0.5... i.e. a fair game


2. Assume class size is 27 instead of 35.  
a)With the same bet ($1 to $1) what is the probability Dr. J wins the bet now.  
b)If 20 of the 27 students make the bet, what is Dr. J's expected profit/loss?

3. What size class (how many students) makes the $1 againt $1 a fair bet?  I.e. find the sample size that makes the mean probability close to 0.5


```{r setup, include=FALSE}
set.seed(123)  # Set seed for reproducibility
```

```{r}
# Function to simulate if two or more people share a birthday
same_birthday_prob <- function(class_size, n_sim) {
  count_same_birthday <- 0
  for (i in 1:n_sim) {
    birthdays <- sample(1:365, class_size, replace = TRUE)  # Generate random birthdays
    if (length(unique(birthdays)) < class_size) {
      count_same_birthday <- count_same_birthday + 1
    }
  }
  return(count_same_birthday / n_sim)  # Return probability
}

# Simulation settings
class_size <- 35
n_sim <- 1000
n_trials <- 100

# Vector to store the probabilities from each trial
probs <- numeric(n_trials)

# Run the simulation 100 times
for (i in 1:n_trials) {
  probs[i] <- same_birthday_prob(class_size, n_sim)
}

```

## mean and std dev 

```{r, echo=F}
# Print mean, std of probabilities

mean_probs <- mean(probs)
sd_probs <- sd(probs)

cat("Mean Probability:", mean_probs, "\n")
cat("Standard Deviation of Probabilities:", sd_probs, "\n")

```

```{r}
hist(probs, main = "Histogram of Probabilities", xlab = "Probability", col = "lightblue", border = "black")

```




```{r}
# Calculate 95% confidence interval
error_margin <- qt(0.975, df=n_trials-1) * sd_probs / sqrt(n_trials)
conf_interval <- c(mean_probs - error_margin, mean_probs + error_margin)

cat("95% Confidence Interval for the Mean Probability: [", conf_interval[1], ",", conf_interval[2], "]\n")

```


get answers for a smaller class 

```{r}
# Simulation settings
class_size <- 26
n_sim <- 1000
n_trials <- 100

# Vector to store the probabilities from each trial
probs <- numeric(n_trials)

# Run the simulation 100 times
for (i in 1:n_trials) {
  probs[i] <- same_birthday_prob(class_size, n_sim)
}

```

```{r}
hist(probs, main = "Histogram of Probabilities", xlab = "Probability", col = "lightblue", border = "black")

```

```{r, echo=F}
# Print mean, std of probabilities

mean_probs <- mean(probs)
sd_probs <- sd(probs)

cat("Mean Probability:", mean_probs, "\n")
cat("Standard Deviation of Probabilities:", sd_probs, "\n")

```
```{r}
# Calculate 95% confidence interval
error_margin <- qt(0.975, df=n_trials-1) * sd_probs / sqrt(n_trials)
conf_interval <- c(mean_probs - error_margin, mean_probs + error_margin)

cat("95% Confidence Interval for the Mean Probability: [", conf_interval[1], ",", conf_interval[2], "]\n")

```

Find the probability that at least two students will have the ssame birthday. If 26 students in the class take the bet how much is Dr. J expected to win/lose. -He is expected to win. 

How much should Dr. J wager against the students $1 to make this a fair bet?  
Dr J would have to bet 4.6 dollars for it to be a fair bet.  

## Question 2

```{r}
# Function to simulate if two or more people share a birthday
same_birthday_prob <- function(class_size, n_sim) {
  count_same_birthday <- 0
  for (i in 1:n_sim) {
    birthdays <- sample(1:365, class_size, replace = TRUE)  # Generate random birthdays
    if (length(unique(birthdays)) < class_size) {
      count_same_birthday <- count_same_birthday + 1
    }
  }
  return(count_same_birthday / n_sim)  # Return probability
}

# Simulation settings
class_size <- 26
n_sim <- 1000
n_trials <- 100

# Vector to store the probabilities from each trial
probs <- numeric(n_trials)

# Run the simulation 100 times
for (i in 1:n_trials) {
  probs[i] <- same_birthday_prob(class_size, n_sim)
}

```

```{r}
# Simulation settings
class_size <- 27
n_sim <- 1000
n_trials <- 100

# Vector to store the probabilities from each trial
probs <- numeric(n_trials)

# Run the simulation 100 times
for (i in 1:n_trials) {
  probs[i] <- same_birthday_prob(class_size, n_sim)
}

```

```{r}
hist(probs, main = "Histogram of Probabilities", xlab = "Probability", col = "lightblue", border = "black")

```


## Mean and std dev
```{r echo=F}
# Print mean, std of probabilities

mean_probs <- mean(probs)
sd_probs <- sd(probs)

cat("Mean Probability:", mean_probs, "\n")
cat("Standard Deviation of Probabilities:", sd_probs, "\n")

```
```{r}
# Calculate 95% confidence interval
error_margin <- qt(0.975, df=n_trials-1) * sd_probs / sqrt(n_trials)
conf_interval <- c(mean_probs - error_margin, mean_probs + error_margin)

cat("95% Confidence Interval for the Mean Probability: [", conf_interval[1], ",", conf_interval[2], "]\n")

```
## Question 3
```{r}
# Function to simulate if two or more people share a birthday
same_birthday_prob <- function(class_size, n_sim) {
  count_same_birthday <- 0
  for (i in 1:n_sim) {
    birthdays <- sample(1:365, class_size, replace = TRUE)  # Generate random birthdays
    if (length(unique(birthdays)) < class_size) {
      count_same_birthday <- count_same_birthday + 1
    }
  }
  return(count_same_birthday / n_sim)  # Return probability
}

# Simulation settings
class_size <- 26
n_sim <- 1000
n_trials <- 100

# Vector to store the probabilities from each trial
probs <- numeric(n_trials)

# Run the simulation 100 times
for (i in 1:n_trials) {
  probs[i] <- same_birthday_prob(class_size, n_sim)
}

```

```{r}
# Simulation settings
class_size <- 23
n_sim <- 1000
n_trials <- 100

# Vector to store the probabilities from each trial
probs <- numeric(n_trials)

# Run the simulation 100 times
for (i in 1:n_trials) {
  probs[i] <- same_birthday_prob(class_size, n_sim)
}

```

```{r}
hist(probs, main = "Histogram of Probabilities", xlab = "Probability", col = "lightblue", border = "black")

```
Mean and std dev

```{r echo=F}
# Print mean, std of probabilities

mean_probs <- mean(probs)
sd_probs <- sd(probs)

cat("Mean Probability:", mean_probs, "\n")
cat("Standard Deviation of Probabilities:", sd_probs, "\n")

```

```{r}
# Calculate 95% confidence interval
error_margin <- qt(0.975, df=n_trials-1) * sd_probs / sqrt(n_trials)
conf_interval <- c(mean_probs - error_margin, mean_probs + error_margin)

cat("95% Confidence Interval for the Mean Probability: [", conf_interval[1], ",", conf_interval[2], "]\n")

```
In order for it to be a fair bet the class would need to have 23 students.

### Explanation:

- **`same_birthday_prob` function**: Simulates whether at least two people in a group of `class_size` share a birthday, repeated `n_sim` times. It returns the proportion of simulations where a birthday match occurs.
- **`probs` vector**: Stores the results of running the birthday problem simulation 100 times.
- **Mean, standard deviation, and histogram**: Summarizes and visualizes the distribution of probabilities.
- **Confidence interval**: Calculates a 95% confidence interval for the mean of the simulated probabilities.
